A general and efficient representation of Ancestral Recombination Graphs

### 2022-09-29 paper.tex
Rejig and condense text about "big" ARGs etc
I have tried to make the link between sample-resolving big ARGs, and (more importantly) forwards simulations, more explicit, as well as giving a tee-up into the simplification section.

There are two distinct annotation schemes possible, and the distinction is subtle
and important. The first scheme is to use the annotations to describe the complete
set of genomic intervals directly inherited by the descendant node from the ancestor
node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
two gARGs with alternative edge annotation schemes.
(A) A conventional Griffiths eARG. Square nodes represent events, with
recombinations (in red) annotated with the position of a single breakpoint
along the 7-unit long chromosome.
(B) A gARG with direct-inheritance edge annotations. Oval nodes represent
genomes and annotations reflect ``local'' transmission of all genetic material
between parent and child. Edges that do not span the full genome are visually
highlighted, showing how breakpoint positions can be encoded.
(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
the relative importance of different transmission pathways.
}
\end{figure}

ARGs are inherently a retrospective structure defined in terms
of a set of sample nodes (usually the leaves), and,
as we discuss in section XXX, storing sample-resolved ancestry
intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
useful than storing the parent-child intervals recording direct inheritance
(Fig~\ref{fig-ancestry-resolution}B).
Indeed, storing direct inheritance intervals for edges is really only of interest because of
its correspondence with the classical approach of associating
breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
The eARG encoding is simply a more limited form, in which the only types of direct
inheritance we can have are the entire genome, or the intervals
at either side of a recombination breakpoint.

Note that this example is a sample from the so-called \emph{``big'' ARG process}
(see Appendix XXX) and has several features that we are unlikely to simulate or
infer in practice. For example, by definition the samples contain no genetic information about
nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
Likewise, is is unlikely we would have information to infer the ``grand MRCA''
\textsf{q} as all regions of the genome have already reached a
common ancestor separately (in nodes \textsf{k} and \textsf{p}).

@@

Given the direct-inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}B),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the sample-resolved material for a node is the
union of the sample-resolved intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their direct-inheritance intervals
and the just-computed sample-resolved intervals for the node.
For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
joining \textsf{i} to its ancestor \textsf{k} then carries
material on $(0, 2]$ since although \textsf{i} directly inherits interval
$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
ancestral to the sample. This process is directly analogous to Hudson's
simulation algorithm
(Appendix XXX).

These alternatives of recording the direct inheritance of genetic material from
parent to child vs the propagation of genomic intervals ancestral to the sample
are analogous to running forwards-in-time vs backwards-in-time simulations.
Indeed, the process of taking a graph with edges annotated with direct parent-child
inheritance (Fig~\ref{fig-ancestry-resolution}B) and producing the equivalent
``ancestry resolved'' graph in which only those genomic intervals that are
ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}C)
is the same key algorithm underlying recent advances in forwards-in-time simulation.
\cite{kelleher2018efficient} developed a method in which forwards-in-time
simulations are performed by recording all parent-child genetic relationships
(i.e.~annotating the edges in the simulated pedigree with the genomic
intervals inherited by a child from a parent) and periodically ``simplifying''
this structure to retain only ancestral relationships that are relevant
to the ancestry of the current population. In many situations, using this approach to
record the full genetic ancestry of the current population is orders of magnitude faster
than the alternative of tracking
sequences~\citep{kelleher2018efficient,haller2018tree}.

>>>>

There are two distinct annotation schemes possible, and the distinction is subtle
and important. The first scheme is to use the annotations to describe the complete
set of genomic intervals directly inherited by the descendant node from the ancestor
node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
As we discuss in section XXX, this second scheme has many advantages
in terms of efficiency.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
two gARGs with alternative edge annotation schemes.
(A) A conventional Griffiths eARG. Square nodes represent events, with
each recombination (red) annotated with the position of a single breakpoint
along the 7-unit-long chromosome.
(B) A gARG with direct-inheritance edge annotations. Oval nodes represent
genomes and annotations reflect ``local'' transmission of all genetic material
between parent and child. Edges that do not span the full genome are visually
highlighted, showing how breakpoint positions can be encoded.
(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}; very few edges then span the entire genome.
The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
the relative importance of different transmission pathways to the samples.
}
\end{figure}

%ARGs are inherently a retrospective structure defined in terms
%of a set of sample nodes (usually the leaves), and,
%as we discuss in section XXX, storing sample-resolved ancestry
%intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
%useful than storing the parent-child intervals recording direct inheritance
%(Fig~\ref{fig-ancestry-resolution}B).
%Indeed, storing direct inheritance intervals for edges is really only of interest because of
%its correspondence with the classical approach of associating
%breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
%The eARG encoding is simply a more limited form, in which the only types of direct
%inheritance we can have are the entire genome, or the intervals
%at either side of a recombination breakpoint.

%Note that this example is a sample from the so-called \emph{``big'' ARG process}
%(see Appendix XXX) and has several features that we are unlikely to simulate or
%infer in practice. For example, by definition the samples contain no genetic information about
%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
%\textsf{q} as all regions of the genome have already reached a
%common ancestor separately (in nodes \textsf{k} and \textsf{p}).

@@

Given the direct-inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}B),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the sample-resolved material for a node is the
union of the sample-resolved intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their direct-inheritance intervals
and the just-computed sample-resolved intervals for the node.
For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
joining \textsf{i} to its ancestor \textsf{k} then carries
material on $(0, 2]$ since although \textsf{i} directly inherits interval
$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
ancestral to the sample. For each node, we also keep track of the number of
descendant samples over each genomic interval, and stop tracking that
interval once the lineages from all samples have coalesced. For example,
the edge from \textsf{k} to \textsf{n} omits the region $(0,2]$ because it has
already fully coalesced (indicated by a fully black genomic region).
This process is directly analogous to Hudson's simulation algorithm
(Appendix XXX).

In practice, the backwards-in-time Griffiths process that produces
ARGs such as that shown in Figs~{fig-ancestry-resolution}A~and~B
is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
dominated by nodes such as \textsf{g} and \textsf{j} which contain information
about inheritance in non-ancestral material, as well as nodes such as
\textsf{o} and \textsf{q} which track details of ancestry even once the
local genomic region has fully coalesced. Instead, simulating
ARGs from the coalescent with recombination is usually done by
running Hudson's algorithm to produce a sample-resolved ARG directly.
Moreover, when inferring ARGs from genetic data, there is no information
in the sample set that can be used to resolve these nodes, so that they
are essentially unknowable.

Nevertheless, there is a practical use for converting a direct-inheritance
gARG to a sample resolved version. It is the key algorithm underlying recent
advances in forwards-in-time simulation.
\cite{kelleher2018efficient} developed a method in which forwards-in-time
simulations are performed by recording all parent-child genetic relationships
(i.e.~annotating the edges in the simulated pedigree with the genomic
intervals inherited by a child from a parent) and periodically ``simplifying''
this structure to retain only ancestral relationships that are relevant
to the ancestry of the current population. In many situations, using this approach to
record the full genetic ancestry of the current population is orders of magnitude faster
than the alternative of tracking neutral genetic variation directly~\citep{kelleher2018efficient,haller2018tree}.

It is worth noting that nodes \textsf{g},  \textsf{j}, \textsf{o} and \textsf{q}
in Fig~\ref{fig-ancestry-resolution}C are redundant in that they can
be removed without reducing our knowledge of the genetic
ancestry of the samples. Removing nodes,
both redundant ones and others, is part of a process we call
``simplification''.
